---
title: "Freeze/Thaw Analyses"
author: "McCaide Wooten"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
---

# Setup

## Packages, Paths, and Constants

```{r}
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(mgcv)
library(gratia)
library(MASS)
library(systemfonts)
library(ragg)
library(forcats)
library(patchwork)
library(beepr)
select <- dplyr::select

theme_set(theme_bw(base_size = 11))

DATA_PATH <- "C:/INSERT FILE PATH HERE/Freeze-Thaw Working Dataset.xlsx"

## Detection limits back-calculated from the drop-plate method
## (6 x 10 uL drops; smallest quantifiable result = 1 colony across 6 drops
##  = (1/6)/0.010 mL = 16.67 CFU/mL, read from the undiluted drop).
##   broth: 16.67 CFU/mL
##   soil : 16.67 x 10 for the 1:10 soil:PBS slurry step -> 166.7 CFU/g
LOD <- c(soil = 166.6667, broth = 16.6667)

K_GLOBAL  <- 10    # basis size for the global smooth of time (13 unique times, so k <= 13)
K_GROUP   <- 8     # basis size for group-level smooths
NSIM      <- 4000  # posterior draws for contrasts / AUC / crossing times
TMAX      <- 100   # trial length, days
GRID_STEP <- 0.5   # spacing of the posterior evaluation grid, days
P_CROSSED <- 0.90  # min. share of draws that must cross before a TTT is reported
SEED      <- 2026

## Analytical threshold: a 4 log10 (99.99%) reduction from baseline. Applied to
## all relative-scale models as the reference level for AUC, the day-100 margin, and the time to
## 99.99% reduction.
THR_REL <- -4

## The absolute-scale broth sensitivity models predict log10 CFU/mL. Those models are
## floored at 0 instead, i.e. 1 CFU/mL.
THR_BROTH_ABS <- 0
```

## Shared Functions

```{r}
## Scale parameter: cnorm stores the SD as exp(theta); getTheta(TRUE) returns
## it on the SD scale. Falls back to sqrt(sig2) for Gaussian fits.
get_sigma <- function(m) {
  if (!is.null(m$family$getTheta)) {
    th <- try(m$family$getTheta(TRUE), silent = TRUE)
    if (!inherits(th, "try-error") && length(th) >= 1) return(as.numeric(th[1]))
  }
  sqrt(m$sig2)
}

## ---- mgcv version guard and censored-Gaussian AIC check -----------------
## mgcv 1.9-4's cnorm()$aic misclassifies uncensored observations <= 0 as
## left-censored (`yat <= 0` instead of `yat == -Inf`) and drops log(sigma) from
## the uncensored density. This analysis is pinned to version 1.9-3.
if (packageVersion("mgcv") == "1.9-4")
  stop("mgcv 1.9-4 reports incorrect AIC for cnorm models; install 1.9-3.")

## Direct censored-Gaussian log-likelihood (prior weights scale the SD as
## sigma / sqrt(w)). Used only to confirm AIC() is right, so a future mgcv
## update can't silently change the model comparison.
cnorm_loglik <- function(m) {
  yy  <- model.response(model.frame(m))
  y   <- yy[, 1]; yat <- yy[, 2]
  mu  <- as.vector(m$linear.predictors)
  s   <- get_sigma(m) / sqrt(m$prior.weights)
  sum(ifelse(yat == -Inf,
             pnorm(y, mu, s, log.p = TRUE),
             dnorm(y, mu, s, log = TRUE)))
}

cnorm_aic_ok <- function(m, tol = 1e-6) {
  if (!grepl("^cnorm", m$family$family)) return(TRUE)
  abs(as.numeric(logLik(m)) - cnorm_loglik(m)) < tol
}

## Brier and logarithmic scores for the model-implied non-detection pattern.
score_detect <- function(d, label) {
  d %>%
    group_by(model) %>%
    summarise(subset   = label,
              n        = n(),
              observed = sum(censored),
              expected = round(sum(p_below), 1),
              brier    = round(mean((p_below - as.numeric(censored))^2), 4),
              logscore = round(-mean(ifelse(censored,
                                            log(pmax(p_below, 1e-12)),
                                            log(pmax(1 - p_below, 1e-12)))), 4),
              .groups  = "drop")
}

# ---- Fitted values over a grid --------------------------
pred_grid <- function(m, nd, label, tag = "model") {
  p <- predict(m, newdata = nd, se.fit = TRUE)
  out <- transform(nd, fit = as.vector(p$fit), se = as.vector(p$se.fit))
  out[[tag]] <- label
  out
}

# ---- Posterior simulation -------------------------------
## Draws of the linear predictor over a prediction grid. vcov is corrected for
## smoothing-parameter uncertainty (unconditional = TRUE).
post_draws <- function(m, nd, nsim = NSIM) {
  Xp <- predict(m, newdata = nd, type = "lpmatrix")
  V  <- vcov(m, unconditional = TRUE)
  b  <- MASS::mvrnorm(nsim, coef(m), V)
  Xp %*% t(b)
}

## Per-draw summary statistics. Every one takes the same (y, t, thr) signature
## so a single dispatcher can drive them all; statistics with no threshold
## simply ignore thr.
first_cross <- function(y, t, thresh) {
  below <- y <= thresh
  if (!any(below)) return(NA_real_)
  t[which.max(below)]
}

auc_floored <- function(y, t, thresh) {
  z <- pmax(y - thresh, 0)
  sum(diff(t) * (head(z, -1) + tail(z, -1)) / 2)
}

stat_auc     <- function(y, t, thr) auc_floored(y, t, thr)   # soil persistence
stat_margin  <- function(y, t, thr) y[length(y)] - thr       # day-100 margin
stat_ttt     <- function(y, t, thr) first_cross(y, t, thr)   # time to threshold

## Dispatcher. Applies a per-draw statistic across the levels of any grouping
## factor. thr is either a single value (THR_REL for every relative-scale model,
## THR_BROTH_ABS for the absolute-scale broth sensitivity) or a vector named by
## level, giving each level its own threshold (own-floor sensitivity).
by_group_stat <- function(D, nd, gvar, fun, thr = NA_real_) {
  g    <- nd[[gvar]]
  levs <- levels(g)
  if (length(thr) > 1 && !all(levs %in% names(thr)))
    stop("thr has more than one value but is not named for every level of ", gvar)
  bind_rows(lapply(levs, function(s) {
    idx   <- which(g == s)
    thr_s <- if (length(thr) > 1) unname(thr[[s]]) else unname(thr)
    v     <- apply(D[idx, , drop = FALSE], 2, fun,
                   t = nd$time[idx], thr = thr_s)
    tibble::tibble(level = s, draw = seq_along(v), value = as.numeric(v))
  })) %>%
    mutate(level = factor(level, levels = levs)) %>%
    rename(!!gvar := level)
}

## The three reported metrics, computed together for one set of posterior
## draws. Every model gets the same set.
metric_set <- function(D, nd, gvar, thr,
                       ttt_label = "time to 99.99% reduction") bind_rows(
  by_group_stat(D, nd, gvar, stat_auc,    thr) %>% mutate(stat = "AUC"),
  by_group_stat(D, nd, gvar, stat_margin, thr) %>% mutate(stat = "day-100 margin"),
  by_group_stat(D, nd, gvar, stat_ttt,    thr) %>% mutate(stat = ttt_label)
)

## Posterior summary of a per-draw metric, for any grouping variable.
summarise_stat <- function(d, gvar) {
  d %>%
    group_by(across(any_of("fit")), .data[[gvar]]) %>%
    summarise(p_above   = round(mean(value > 0, na.rm = TRUE), 3),
              p_crossed = round(mean(!is.na(value)), 3),
              median = median(value, na.rm = TRUE),
              q25 = quantile(value, 0.25, na.rm = TRUE),
              q75 = quantile(value, 0.75, na.rm = TRUE),
              lwr = quantile(value, 0.025, na.rm = TRUE),
              upr = quantile(value, 0.975, na.rm = TRUE),
              skew = (quantile(value, 0.975, na.rm = TRUE) - median(value, na.rm = TRUE)) /
                     (median(value, na.rm = TRUE) - quantile(value, 0.025, na.rm = TRUE)),
              .groups = "drop") %>%
    mutate(across(c(median, q25, q75, lwr, upr, skew), ~ round(.x, 2)))
}

## Differences taken within draw. pairs = NULL gives every pairwise
## combination; supply a list of pairs for the within-strain regime comparisons.
contrast_draws <- function(d, gvar, label, pairs = NULL) {
  wide <- d %>%
    select(all_of(c(gvar, "draw", "value"))) %>%
    tidyr::pivot_wider(names_from = all_of(gvar), values_from = value)
  levs <- setdiff(names(wide), "draw")
  if (is.null(pairs)) pairs <- utils::combn(levs, 2, simplify = FALSE)
  bind_rows(lapply(pairs, function(p) {
    if (!all(p %in% names(wide))) return(NULL)
    dif <- wide[[p[1]]] - wide[[p[2]]]
    tibble::tibble(
      stat     = label,
      contrast = paste0(p[1], " - ", p[2]),
      n_usable = sum(!is.na(dif)),
      median   = median(dif, na.rm = TRUE),
      lwr      = quantile(dif, 0.025, na.rm = TRUE),
      upr      = quantile(dif, 0.975, na.rm = TRUE),
      p_gt0    = mean(dif > 0, na.rm = TRUE)
    )
  })) %>%
    mutate(excludes_0 = (lwr > 0 | upr < 0),
           across(c(median, lwr, upr), ~ round(.x, 2)))
}

## Do two contrast tables reach the same conclusion? Used by every sensitivity
## analysis (absolute-scale broth, GI penalty order, own-floor thresholds).
sens_agree <- function(ca, cb, label, names_ab = c("a", "b")) {
  bind_rows(mutate(ca, fit = names_ab[1]), mutate(cb, fit = names_ab[2])) %>%
    select(fit, contrast, excludes_0) %>%
    tidyr::pivot_wider(names_from = fit, values_from = excludes_0) %>%
    mutate(stat  = label,
           agree = .data[[names_ab[1]]] == .data[[names_ab[2]]])
}

## Contrast agreement across every metric in two contrast tables.
sens_agree_all <- function(ca, cb, names_ab) {
  bind_rows(lapply(intersect(unique(ca$stat), unique(cb$stat)), function(st)
    sens_agree(filter(ca, stat == st), filter(cb, stat == st), st, names_ab)))
}

## ---- Table-ready strings for supplementary tables -----------------------
## "201 (174, 232)" for estimates; for unreportable times "NA; 80.4%" (share of
## draws that crossed), so it can't be misread as a range.
fmt_est <- function(med, lwr, upr, digits = 3) {
  f <- function(x) trimws(formatC(signif(x, digits), format = "fg", digits = digits))
  ifelse(is.na(med), NA_character_, paste0(f(med), " (", f(lwr), ", ", f(upr), ")"))
}
fmt_ttr <- function(med, lwr, upr, p_crossed, reportable) {
  ifelse(reportable, fmt_est(med, lwr, upr),
         paste0("NA; ", formatC(100 * p_crossed, format = "f", digits = 1), "%"))
}
```

# Data Handling

## Load and Type

```{r}
ft_raw <- read_excel(DATA_PATH, sheet = "Data")

ft <- ft_raw %>%
  mutate(
    strain    = factor(strain, levels = 1:6),
    medium    = factor(medium, levels = c("broth", "soil")),
    group     = factor(group,  levels = c("exp", "plus10C", "minus10C")),
    replicate = factor(replicate),
    series    = interaction(strain, medium, group, drop = TRUE, sep = "_"),
    lod       = unname(LOD[as.character(medium)]),
    censored  = concentration == 0
  )
```

## Strain Metadata

```{r}
strain_meta <- tibble::tribble(
  ~strain, ~species,           ~clade,       ~host,        ~is_AC,
  "1",     "E. rhusiopathiae", "2",          "Poultry",    FALSE,
  "2",     "E. rhusiopathiae", "2",          "Arctic fox", FALSE,
  "3",     "E. rhusiopathiae", "3 (non-AC)", "Swine",      FALSE,
  "4",     "E. rhusiopathiae", "3 (non-AC)", "Muskox",     FALSE,
  "5",     "E. rhusiopathiae", "3 (AC)",     "Muskox",     TRUE,
  "6",     "E. qupirruqcola",  NA,           "Larvae",     FALSE
) %>%
  mutate(strain = factor(strain, levels = 1:6))

ft <- left_join(ft, strain_meta, by = "strain")

## Facet/axis labels used by every diagnostic plot below.
strain_labs <- with(
  strain_meta,
  setNames(
    paste0(strain, ": ", host, " — ",
           ifelse(is.na(clade), "E. qupirruqcola", paste0("clade ", clade))),
    as.character(strain)
  )
)
```

## Culture Stopping Rule

```{r}
## Protocol: culturing stopped once a strain/medium was negative at two
## successive timepoints.
stop_rule <- tibble::tibble(
  strain     = factor("3", levels = 1:6),
  medium     = factor("soil", levels = c("broth", "soil")),
  group      = factor("exp", levels = c("exp", "plus10C", "minus10C")),
  stop_after = 30
)

ft <- ft %>%
  left_join(stop_rule, by = c("strain", "medium", "group")) %>%
  mutate(stop_after = ifelse(is.na(stop_after), Inf, stop_after),
         measured   = time <= stop_after) %>%
  select(-stop_after)
```

## Baselines and Response Construction

```{r}
## Baseline = mean log10 concentration of the three day-0 replicates, per
## strain/medium/regime series. The response is the log10 fold change from that
## baseline. Non-detections take the LOD in y_obs (used for plotting and for
## the Gaussian broth fits) and -Inf in y_cens, which is what cnorm() reads as
## "left censored at y_obs". y_abs is the same quantity on the absolute scale,
## used by the two absolute-scale broth models.
stopifnot(all(ft$concentration[ft$time == 0] > 0))

baselines <- ft %>%
  filter(time == 0) %>%
  group_by(series) %>%
  summarise(log10_baseline = mean(log10(concentration)), .groups = "drop")

ft <- left_join(ft, baselines, by = "series") %>%
  mutate(
    y_abs  = log10(ifelse(censored, lod, concentration)),
    y_obs  = y_abs - log10_baseline,
    y_cens = ifelse(censored, -Inf, y_obs)
  )

# ---- Analysis subsets ------------------------------------------------------
ft_exp   <- filter(ft, group == "exp") %>% droplevels()
ft_broth <- filter(ft_exp, medium == "broth") %>% droplevels()
ft_soil  <- filter(ft_exp, medium == "soil")  %>% droplevels()
```

## Detection Limits vs Analytical Threshold

```{r}
lod_rel <- ft %>%
  distinct(series, strain, medium, group, lod, log10_baseline) %>%
  mutate(y_lod = log10(lod) - log10_baseline)

## Which series could resolve a 4-log decline directly, and how far below its
## own detection floor each of the others has to be extrapolated to get there.
lod_rel %>%
  filter(medium == "soil") %>%
  transmute(series,
            own_floor       = round(y_lod, 3),
            resolvable_at_4 = y_lod <= THR_REL,
            extrapolated_log = round(pmax(y_lod - THR_REL, 0), 3),
            truncated_log    = round(pmax(THR_REL - y_lod, 0), 3)) %>%
  arrange(desc(own_floor))

```

## Data Checks and Cell Summaries

```{r}
stopifnot(nrow(ft) == 780)
stopifnot(all(count(ft, series, time)$n == 3))
stopifnot(all(is.finite(ft$y_obs)))
stopifnot(sum(!ft$measured) == 21)
stopifnot(all(ft$censored[!ft$measured]))

ft %>%
  group_by(medium, group) %>%
  summarise(n = n(),
            n_censored = sum(censored),
            n_imputed  = sum(!measured),
            min_y = min(y_obs), max_y = max(y_obs), .groups = "drop")

## Replicate-level spread
cell_stats <- ft_exp %>%
  group_by(medium, strain, time) %>%
  summarise(mean_y = mean(y_obs), sd_y = sd(y_obs),
            n_cens = sum(censored), .groups = "drop")

## Censoring map: how many of the 3 soil replicates were non-detects
ft_soil %>%
  group_by(strain, time) %>%
  summarise(n_cens = sum(censored), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = strain, values_from = n_cens,
                     names_prefix = "S")
```

## Orientation Plots

```{r}
# ---- All experimental trajectories, both media -----------------------------
## Dotted red = each series' own analytical floor; solid red = THR_REL (-4).
ggplot(ft_exp, aes(x = time, y = y_obs)) +
  geom_hline(yintercept = 0, linetype = 2, colour = "grey60") +
  geom_hline(data = tibble::tibble(medium = factor("soil", levels = levels(ft$medium))),
             aes(yintercept = THR_REL), colour = "firebrick", linewidth = 0.5) +
  geom_segment(data = filter(lod_rel, group == "exp"),
               aes(x = 0, xend = TMAX, y = y_lod, yend = y_lod),
               linetype = 3, colour = "firebrick", inherit.aes = FALSE) +
  geom_point(aes(shape = censored), alpha = 0.55, size = 1.4) +
  stat_summary(fun = mean, geom = "line", aes(group = 1), linewidth = 0.7) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 2),
                     labels = c("detected", "below LOD")) +
  facet_grid(medium ~ strain, labeller = labeller(strain = strain_labs)) +
  labs(x = "Day (= freeze/thaw cycles)",
       y = expression(log[10]~fold~change~from~baseline), shape = NULL) +
  theme(legend.position = "bottom")

# ---- Replicate spread vs time and vs level ---------------------------------
ggplot(cell_stats, aes(time, sd_y, colour = strain)) +
  geom_point(aes(size = n_cens > 0)) +
  geom_line(aes(group = strain), alpha = 0.6) +
  scale_size_manual(values = c(`FALSE` = 1.3, `TRUE` = 2.6), guide = "none") +
  facet_wrap(~ medium) +
  labs(x = "Day", y = "SD of the 3 replicates (log10 units)")

ggplot(cell_stats, aes(mean_y, sd_y, colour = medium)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 0.7) +
  labs(x = "Cell mean (log10 fold change)", y = "Cell SD (log10 units)")

# ---- Cycling vs constant temperature, strains 4 and 6 ----------------------
ggplot(filter(ft, strain %in% c("4", "6")) %>% droplevels(),
       aes(x = time, y = y_obs, colour = group)) +
  geom_hline(yintercept = 0, linetype = 2, colour = "grey60") +
  geom_point(alpha = 0.5, size = 1.3) +
  stat_summary(fun = mean, geom = "line", aes(group = group), linewidth = 0.8) +
  facet_grid(medium ~ strain, labeller = labeller(strain = strain_labs)) +
  labs(x = "Day", y = expression(log[10]~fold~change~from~baseline),
       colour = NULL) +
  theme(legend.position = "bottom")
```

# Model Structure

## Candidate Structures

```{r}
# ---- The five Pedersen structures, plus m-variants of GI and I --------------
## pc = 0 pins each smooth at zero on day 0; "- 1" removes the intercept so the
## fitted value is zero there too, not just the smooth.
## GI uses a first-derivative (m = 1) penalty on the strain smooths, as in
## Pedersen et al. (2019). GI_m2 and I_m1 complete a 2 x 2 of {global term
## present/absent} x {strain-smooth penalty order}, so AIC can separate the two.
## Note: with pc = 0, an m = 2 strain smooth keeps an unpenalized linear slope
## (edf >= 1); an m = 1 strain smooth can shrink all the way to zero.
hgam_forms <- list(
  G     = ~ s(time, k = K_GLOBAL, m = 2, pc = 0) - 1,

  GS    = ~ s(time, k = K_GLOBAL, m = 2, pc = 0) +
            s(time, strain, bs = "fs", k = K_GROUP, m = 2, pc = 0) - 1,

  GI    = ~ s(time, k = K_GLOBAL, m = 2, pc = 0) +
            s(time, by = strain, k = K_GROUP, m = 1, pc = 0) - 1,

  GI_m2 = ~ s(time, k = K_GLOBAL, m = 2, pc = 0) +
            s(time, by = strain, k = K_GROUP, m = 2, pc = 0) - 1,

  S     = ~ s(time, strain, bs = "fs", k = K_GROUP, m = 2, pc = 0) - 1,

  I     = ~ s(time, by = strain, k = K_GROUP, m = 2, pc = 0) - 1,

  I_m1  = ~ s(time, by = strain, k = K_GROUP, m = 1, pc = 0) - 1
)

## Fallbacks: the "fs" basis construction does not accept pc, so GS and S
## revert to the unconstrained form, which needs its intercept back.
hgam_alt <- list(
  GS = ~ s(time, k = K_GLOBAL, m = 2, pc = 0) +
         s(time, strain, bs = "fs", k = K_GROUP, m = 2),
  S  = ~ s(time, strain, bs = "fs", k = K_GROUP, m = 2)
)

## Verify the formulas before committing to any fitting
lapply(hgam_forms, update, y_obs ~ .)
```

## Fit Both Media

```{r}
## Broth: Gaussian (no non-detections occurred in broth).
## Soil : censored Gaussian (cnorm), left censored at each series' own LOD.
fit_one <- function(f, lhs, data, fam_fn, f_alt = NULL) {
  out <- tryCatch(
    list(fit  = gam(update(f, lhs), data = data, family = fam_fn(),
                    method = "REML"),
         spec = "pc", note = NA_character_),
    error = function(e) list(fit = NULL, spec = NA_character_,
                             note = conditionMessage(e))
  )
  if (is.null(out$fit) && !is.null(f_alt)) {
    first_note <- out$note
    out <- tryCatch(
      list(fit  = gam(update(f_alt, lhs), data = data, family = fam_fn(),
                      method = "REML"),
           spec = "no_pc", note = first_note),
      error = function(e) list(fit = NULL, spec = NA_character_,
                               note = conditionMessage(e))
    )
  }
  out
}

fit_hgam_set <- function(data, lhs, fam_fn) {
  setNames(lapply(names(hgam_forms), function(nm) {
    fit_one(hgam_forms[[nm]], lhs, data, fam_fn, hgam_alt[[nm]])
  }), names(hgam_forms))
}

m_broth <- fit_hgam_set(ft_broth, y_obs ~ .,                gaussian)
m_soil  <- fit_hgam_set(ft_soil,  cbind(y_obs, y_cens) ~ .,  cnorm)
```

# Model Selection, Verification, and Variations

## AIC Comparison and Candidate Curves

```{r}
compare_hgam <- function(mods, label) {
  ok <- !vapply(mods, function(x) is.null(x$fit), logical(1))
  for (nm in names(mods)[!ok])
    message(label, " / ", nm, " failed: ", mods[[nm]]$note)
  for (nm in names(mods)[ok])
    if (identical(mods[[nm]]$spec, "no_pc"))
      message(label, " / ", nm, " fell back to no-pc: ", mods[[nm]]$note)
  mods <- mods[ok]

  tibble::tibble(
    medium   = label,
    model    = names(mods),
    spec     = vapply(mods, function(x) x$spec, character(1)),
    n_sp     = vapply(mods, function(x) length(x$fit$sp), integer(1)),
    edf      = vapply(mods, function(x) sum(x$fit$edf), numeric(1)),
    dev_expl = vapply(mods, function(x) summary(x$fit)$dev.expl, numeric(1)),
    REML     = vapply(mods, function(x) as.numeric(x$fit$gcv.ubre), numeric(1)),
    logLik   = vapply(mods, function(x) as.numeric(logLik(x$fit)), numeric(1)),
    AIC      = vapply(mods, function(x) AIC(x$fit), numeric(1)),
    aic_ok   = vapply(mods, function(x) cnorm_aic_ok(x$fit), logical(1))
  ) %>%
    arrange(AIC) %>%
    mutate(dAIC = AIC - min(AIC), .after = AIC)
}

aic_tab <- bind_rows(compare_hgam(m_broth, "broth"), compare_hgam(m_soil, "soil"))
stopifnot(all(aic_tab$aic_ok))   # cnorm AIC() matches the direct likelihood
print(aic_tab %>% mutate(across(c(edf, dev_expl, REML, logLik, AIC, dAIC), ~ round(.x, 2))),
      n = Inf, width = Inf)

## The 2 x 2 that separates the global term from penalty order
aic_tab %>%
  filter(model %in% c("GI", "GI_m2", "I_m1", "I")) %>%
  mutate(global  = ifelse(grepl("^GI", model), "global + strain", "strain only"),
         penalty = ifelse(model %in% c("GI", "I_m1"), "m = 1", "m = 2")) %>%
  select(medium, global, penalty, AIC) %>%
  mutate(AIC = round(AIC, 1)) %>%
  tidyr::pivot_wider(names_from = penalty, values_from = AIC) %>%
  select(medium, global, `m = 1`, `m = 2`)

# ---- Do the point constraints behave? Fitted values at day 0 and day 100 ---
endpoint_check <- function(mods, data, label, at) {
  nd <- expand.grid(time = at,
                    strain = factor(levels(data$strain),
                                    levels = levels(data$strain)))
  bind_rows(lapply(names(mods), function(nm) {
    if (is.null(mods[[nm]]$fit)) return(NULL)
    tibble::tibble(medium = label, model = nm, strain = nd$strain, day = at,
                   fit = as.vector(predict(mods[[nm]]$fit, newdata = nd)))
  })) %>%
    tidyr::pivot_wider(names_from = strain, values_from = fit,
                       names_prefix = "S")
}

endpoints <- bind_rows(
  endpoint_check(m_broth, ft_broth, "broth", 0),
  endpoint_check(m_soil,  ft_soil,  "soil",  0),
  endpoint_check(m_broth, ft_broth, "broth", TMAX),
  endpoint_check(m_soil,  ft_soil,  "soil",  TMAX)
)

## Day 0: should be 0 to machine precision for every "pc" model; GS and S
## (no_pc fallbacks) carry their fitted intercept instead.
print(endpoints %>% filter(day == 0) %>%
        mutate(across(starts_with("S"), ~ signif(.x, 2))),
      n = Inf, width = Inf)

## Day 100: where each candidate ends up (soil curves can extrapolate far below
## the detection floor here, which is the point of checking).
print(endpoints %>% filter(day == TMAX) %>%
        mutate(across(starts_with("S"), ~ round(.x, 2))),
      n = Inf, width = Inf)

# ---- Candidate curves ------------------------------------------------------
Y_WINDOW <- c(-6, 1)

newd <- expand.grid(
  time   = seq(0, TMAX, length.out = 200),
  strain = factor(levels(ft_soil$strain), levels = levels(ft_soil$strain))
)

pred_set <- function(mods) {
  bind_rows(lapply(names(mods), function(nm) {
    if (is.null(mods[[nm]]$fit)) return(NULL)
    p <- predict(mods[[nm]]$fit, newdata = newd, se.fit = TRUE)
    transform(newd, fit = as.vector(p$fit), se = as.vector(p$se.fit),
              model = factor(nm, levels = names(hgam_forms)))
  }))
}

ggplot(pred_set(m_soil), aes(time, fit, colour = model)) +
  geom_hline(yintercept = 0, linetype = 2, colour = "grey60") +
  geom_hline(yintercept = THR_REL, colour = "firebrick", linewidth = 0.5) +
  geom_point(data = ft_soil, aes(time, y_obs, shape = censored),
             inherit.aes = FALSE, alpha = 0.35, size = 1.1) +
  geom_line(linewidth = 0.75) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 2), guide = "none") +
  coord_cartesian(ylim = Y_WINDOW) +
  facet_wrap(~ strain, labeller = labeller(strain = strain_labs), ncol = 3) +
  labs(x = "Day", y = expression(log[10]~fold~change~from~baseline),
       colour = NULL, title = "Soil",
       caption = "y-axis clipped at -6; some fitted curves extend below") +
  theme(legend.position = "bottom")

ggplot(pred_set(m_broth), aes(time, fit, colour = model)) +
  geom_hline(yintercept = 0, linetype = 2, colour = "grey60") +
  geom_point(data = ft_broth, aes(time, y_obs),
             inherit.aes = FALSE, alpha = 0.35, size = 1.1) +
  geom_line(linewidth = 0.75) +
  facet_wrap(~ strain, labeller = labeller(strain = strain_labs), ncol = 3) +
  labs(x = "Day", y = expression(log[10]~fold~change~from~baseline),
       colour = NULL, title = "Broth") +
  theme(legend.position = "bottom")

## Named handles for the candidates carried into next section.
gi_broth  <- m_broth$GI$fit
gi2_broth <- m_broth$GI_m2$fit
g_soil    <- m_soil$G$fit
gi_soil   <- m_soil$GI$fit
gi2_soil  <- m_soil$GI_m2$fit
```

## Basis Dimension and Residual Checks

```{r}
tibble::tibble(
  model = c("broth GI", "soil G", "soil GI"),
  sigma = c(get_sigma(gi_broth), get_sigma(g_soil), get_sigma(gi_soil))
)

k.check(gi_broth)
k.check(g_soil)
k.check(gi_soil)

resid_df <- function(m, data, label) {
  data %>%
    mutate(model  = label,
           fitted = as.vector(predict(m)),
           resid  = as.vector(residuals(m, type = "deviance")))
}

rd <- bind_rows(
  resid_df(gi_broth, ft_broth, "broth GI"),
  resid_df(g_soil,   ft_soil,  "soil G"),
  resid_df(gi_soil,  ft_soil,  "soil GI")
)

ggplot(rd, aes(fitted, resid, shape = censored)) +
  geom_hline(yintercept = 0, linetype = 2, colour = "grey60") +
  geom_point(alpha = 0.5, size = 1.3) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 0.6, colour = "firebrick") +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 2),
                     labels = c("detected", "below LOD")) +
  facet_wrap(~ model, scales = "free") +
  labs(x = "Fitted value", y = "Deviance residual", shape = NULL) +
  theme(legend.position = "bottom")

ggplot(rd, aes(time, resid, colour = strain, shape = censored)) +
  geom_hline(yintercept = 0, linetype = 2, colour = "grey60") +
  geom_point(alpha = 0.6, size = 1.2) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 2), guide = "none") +
  facet_wrap(~ model, scales = "free_y") +
  labs(x = "Day", y = "Deviance residual", colour = "Strain")

ggplot(rd, aes(sample = resid)) +
  stat_qq(size = 1) + stat_qq_line(colour = "firebrick") +
  facet_wrap(~ model, scales = "free") +
  labs(x = "Theoretical quantile", y = "Deviance residual")
```

## Detection-Pattern Check (Soil Only)

```{r}
## AIC does not heavily weight censored observations, so the soil candidates
## are also scored on their ability to reproduce the observed non-detections.
detect_check <- function(m, label, wts = NULL) {
  sd_i <- if (is.null(wts)) get_sigma(m) else get_sigma(m) / sqrt(wts)
  ft_soil %>%
    mutate(model   = label,
           mu      = as.vector(predict(m)),
           y_lod   = log10(lod) - log10_baseline,
           p_below = pnorm(y_lod, mean = mu, sd = sd_i))
}

dc <- bind_rows(detect_check(g_soil,   "G"),
                detect_check(gi_soil,  "GI"),
                detect_check(gi2_soil, "GI_m2"))

ggplot(dc %>% group_by(model, strain, time) %>%
         summarise(observed = sum(censored), expected = sum(p_below),
                   .groups = "drop"),
       aes(time)) +
  geom_line(aes(y = observed), colour = "grey30", linewidth = 0.9) +
  geom_point(aes(y = observed), colour = "grey30", size = 1.3) +
  geom_line(aes(y = expected, colour = model), linewidth = 0.8) +
  facet_wrap(~ strain, labeller = labeller(strain = strain_labs), ncol = 3) +
  labs(x = "Day", y = "Replicates below LOD (of 3)",
       colour = "Model predicted", subtitle = "Grey = observed") +
  theme(legend.position = "bottom")

dc %>%
  group_by(strain, model) %>%
  summarise(observed = sum(censored),
            expected = round(sum(p_below), 1), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = model, values_from = expected,
                     names_prefix = "expected_", id_cols = c(strain, observed))

bind_rows(score_detect(dc, "all"), score_detect(filter(dc, measured), "plated only"))

## Calibration
dc %>%
  mutate(bin = cut(p_below, breaks = seq(0, 1, 0.2), include.lowest = TRUE)) %>%
  group_by(model, bin) %>%
  summarise(n = n(), mean_p = round(mean(p_below), 3),
            obs_rate = round(mean(censored), 3), .groups = "drop")
```

> **Selection outcome.** GI — a global smooth plus group-level smooths with
> separate smoothing parameters — is carried forward for both media. The two
> variations below are fitted on that same structure based on agreement in AIC and detection scores.

## Variation 1: Soil GI w/ Time-Varying Residual Scale

```{r}
# ---- Stage 1: estimate how residual scale changes with time ----------------
## Residuals only exist for uncensored observations, so the scale model is
## built from those and extrapolated. Modeled against time rather than fitted
## value because time is observed for every row, including censored ones.
soil_resid <- ft_soil %>%
  mutate(mu = as.vector(predict(gi_soil)), r = y_obs - mu) %>%
  filter(!censored)

nrow(soil_resid)   # 171 of 234

scale_mod <- gam(I(r^2) ~ s(time, k = 5), data = soil_resid,
                 family = Gamma(link = "log"), method = "REML")
summary(scale_mod)

scale_grid <- data.frame(time = seq(0, TMAX, length.out = 200))
scale_grid$sd_hat <- sqrt(as.vector(predict(scale_mod, newdata = scale_grid,
                                            type = "response")))

ggplot(scale_grid, aes(time, sd_hat)) +
  geom_point(data = filter(cell_stats, medium == "soil"),
             aes(time, sd_y), inherit.aes = FALSE, alpha = 0.55, size = 1.4) +
  geom_line(linewidth = 0.9, colour = "firebrick") +
  geom_hline(yintercept = get_sigma(gi_soil), linetype = 2, colour = "grey40") +
  labs(x = "Day", y = "Residual SD (log10 units)",
       subtitle = "Red = fitted scale model; points = observed cell SDs; dashed = single sigma")

# ---- Stage 2: refit GI with inverse-variance prior weights -----------------
ft_soil_w <- ft_soil
ft_soil_w$var_hat <- as.vector(predict(scale_mod, newdata = ft_soil_w,
                                       type = "response"))
ft_soil_w$w <- 1 / ft_soil_w$var_hat
ft_soil_w$w <- ft_soil_w$w / mean(ft_soil_w$w)

ft_soil_w %>%
  group_by(time) %>%
  summarise(w = round(first(w), 3), sd_implied = round(sqrt(first(var_hat)), 3),
            .groups = "drop")

gi_soil_w <- gam(update(hgam_forms$GI, cbind(y_obs, y_cens) ~ .),
                 data = ft_soil_w, family = cnorm(), weights = w,
                 method = "REML")

tibble::tibble(
  fit   = c("GI unweighted", "GI weighted"),
  sigma = c(get_sigma(gi_soil), get_sigma(gi_soil_w)),
  edf   = c(sum(gi_soil$edf),   sum(gi_soil_w$edf))
)

k.check(gi_soil_w)

# ---- Did the weighting flatten the residual spread? ------------------------
ft_soil_w %>%
  mutate(mu = as.vector(predict(gi_soil_w)), r = y_obs - mu,
         rstd = r * sqrt(w)) %>%
  filter(!censored) %>%
  ggplot(aes(time, rstd)) +
  geom_hline(yintercept = 0, linetype = 2, colour = "grey60") +
  geom_point(aes(colour = strain), alpha = 0.65, size = 1.3) +
  geom_smooth(aes(y = abs(rstd)), method = "loess", se = FALSE,
              colour = "firebrick", linewidth = 0.7) +
  labs(x = "Day", y = "Weighted residual",
       subtitle = "Red = loess of |residual|; should now be roughly flat")

# ---- Weighted vs unweighted curves and detection scores -------------------
bind_rows(pred_grid(gi_soil,   newd, "unweighted", tag = "wt"),
          pred_grid(gi_soil_w, newd, "weighted",   tag = "wt")) %>%
  ggplot(aes(time, fit, colour = wt, fill = wt)) +
  geom_hline(yintercept = 0, linetype = 2, colour = "grey60") +
  geom_hline(yintercept = THR_REL, colour = "firebrick", linewidth = 0.5) +
  geom_ribbon(aes(ymin = fit - 2 * se, ymax = fit + 2 * se),
              alpha = 0.18, colour = NA) +
  geom_line(linewidth = 0.8) +
  geom_point(data = ft_soil, aes(time, y_obs, shape = censored),
             inherit.aes = FALSE, alpha = 0.3, size = 1) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 2), guide = "none") +
  coord_cartesian(ylim = Y_WINDOW) +
  facet_wrap(~ strain, labeller = labeller(strain = strain_labs), ncol = 3) +
  labs(x = "Day", y = expression(log[10]~fold~change~from~baseline),
       colour = NULL, fill = NULL, title = "Soil GI: weighted vs unweighted") +
  theme(legend.position = "bottom")

dc2 <- bind_rows(detect_check(gi_soil, "GI unweighted"),
                 detect_check(gi_soil_w, "GI weighted", wts = ft_soil_w$w))

dc2 %>%
  group_by(strain, model) %>%
  summarise(observed = sum(censored),
            expected = round(sum(p_below), 1), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = model, values_from = expected,
                     id_cols = c(strain, observed))

bind_rows(score_detect(dc2, "all"), score_detect(filter(dc2, measured), "plated only"))
```
## Variation 1b: Strain Smooths vs the Global Smooth

```{r}
# ---- Weighted GI_m2, same prior weights as gi_soil_w ----------------------
gi2_soil_w <- gam(update(hgam_forms$GI_m2, cbind(y_obs, y_cens) ~ .),
                  data = ft_soil_w, family = cnorm(), weights = w,
                  method = "REML")
k.check(gi2_soil_w)

# ---- Per-strain edf: penalty order x weighting ----------------------------
edf_terms <- function(m) {
  s <- summary(m)$s.table
  tibble::tibble(term = rownames(s), edf = s[, "edf"])
}
bind_rows(
  edf_terms(gi_soil)    %>% mutate(fit = "GI m=1, unweighted"),
  edf_terms(gi_soil_w)  %>% mutate(fit = "GI m=1, weighted"),
  edf_terms(gi2_soil)   %>% mutate(fit = "GI m=2, unweighted"),
  edf_terms(gi2_soil_w) %>% mutate(fit = "GI m=2, weighted")
) %>%
  mutate(edf = round(edf, 3)) %>%
  tidyr::pivot_wider(names_from = fit, values_from = edf)

tibble::tibble(
  fit = c("GI m=1, weighted", "GI m=2, weighted"),
  AIC = c(AIC(gi_soil_w), AIC(gi2_soil_w)),
  aic_ok = c(cnorm_aic_ok(gi_soil_w), cnorm_aic_ok(gi2_soil_w))
) %>% mutate(dAIC = AIC - min(AIC), .after = AIC)

score_detect(bind_rows(detect_check(gi_soil_w,  "GI m=1, weighted", ft_soil_w$w),
                       detect_check(gi2_soil_w, "GI m=2, weighted", ft_soil_w$w)),
             "all")

# ---- Global-only vs full GI curve, against the raw data -------------------
## The global term comes from the same fit (type = "terms"); with pc = 0 and no
## intercept the full fit is exactly global + strain deviation.
global_vs_group <- function(m, nd, label) {
  pt  <- predict(m, newdata = nd, type = "terms", se.fit = TRUE)
  pf  <- predict(m, newdata = nd, se.fit = TRUE)
  col <- match(paste0("s(time):strain", nd$strain), colnames(pt$fit))
  idx <- cbind(seq_len(nrow(nd)), col)
  tibble::tibble(nd, model = label,
                 global  = pt$fit[, "s(time)"],
                 dev     = pt$fit[idx], dev_se = pt$se.fit[idx],
                 full    = as.vector(pf$fit), full_se = as.vector(pf$se.fit))
}

nd_check <- expand.grid(time = seq(0, TMAX, by = 0.5),
                        strain = factor(levels(ft_soil$strain),
                                        levels = levels(ft_soil$strain)))

gv <- bind_rows(global_vs_group(gi_soil,    nd_check, "GI m=1, unweighted"),
                global_vs_group(gi_soil_w,  nd_check, "GI m=1, weighted"),
                global_vs_group(gi2_soil_w, nd_check, "GI m=2, weighted")) %>%
  mutate(model = factor(model, levels = unique(model)))

ggplot(filter(gv, model != "GI m=2, weighted"), aes(time)) +
  geom_hline(yintercept = THR_REL, colour = "firebrick", linewidth = 0.4) +
  geom_point(data = ft_soil, aes(time, y_obs, shape = censored),
             alpha = 0.5, size = 1.1) +
  geom_ribbon(aes(ymin = full - 2 * full_se, ymax = full + 2 * full_se),
              fill = "steelblue", alpha = 0.15) +
  geom_line(aes(y = global, colour = "Global smooth only"),
            linewidth = 0.7, linetype = 2) +
  geom_line(aes(y = full, colour = "GI (global + strain)"), linewidth = 0.8) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 2),
                     labels = c("detected", "below LOD"), name = NULL) +
  scale_colour_manual(values = c("GI (global + strain)" = "steelblue",
                                 "Global smooth only"   = "grey20"), name = NULL) +
  coord_cartesian(ylim = Y_WINDOW) +
  facet_grid(model ~ strain, labeller = labeller(strain = strain_labs)) +
  labs(x = "Day", y = expression(log[10]~fold~change~from~baseline)) +
  theme(legend.position = "bottom")

## Strain deviation from the global smooth, +/- 2 SE. A band that covers 0
## throughout = the strain's trajectory is not distinguishable from the
## shared trend.
ggplot(gv, aes(time, dev)) +
  geom_hline(yintercept = 0, colour = "grey50") +
  geom_ribbon(aes(ymin = dev - 2 * dev_se, ymax = dev + 2 * dev_se), alpha = 0.25) +
  geom_line() +
  coord_cartesian(ylim = c(-4, 3)) +
  facet_grid(model ~ strain, labeller = labeller(strain = strain_labs)) +
  labs(x = "Day", y = "Strain deviation from global smooth")
```


## Variation 2: Broth on the Absolute Scale

```{r}
## Broth replicates grew to similar plateau concentrations from more variable
## baselines, so fold change undervalues strains that started higher. Same GI
## structure, no point constraint (curves do not share an origin), strain as a
## parametric term.
ft_broth %>%
  group_by(strain) %>%
  summarise(baseline = round(first(log10_baseline), 2),
            peak_obs = round(max(y_abs), 2), .groups = "drop") %>%
  mutate(fold = round(peak_obs - baseline, 2))

m_broth_abs <- gam(
  y_abs ~ strain + s(time, k = K_GLOBAL, m = 2) +
                   s(time, by = strain, k = K_GROUP, m = 1),
  data = ft_broth, family = gaussian(), method = "REML"
)

k.check(m_broth_abs)
summary(m_broth_abs)$dev.expl
```

# Derived Quantities and Model Inference

## Prediction Grids and Posterior Draws

```{r}
set.seed(SEED)

t_grid <- seq(0, TMAX, by = GRID_STEP)

nd_soil  <- expand.grid(time = t_grid,
                        strain = factor(levels(ft_soil$strain),
                                        levels = levels(ft_soil$strain)))
nd_broth <- expand.grid(time = t_grid,
                        strain = factor(levels(ft_broth$strain),
                                        levels = levels(ft_broth$strain)))

## The weighted soil fit is the one retained for interpretation (Variation 1),
## so the unweighted fit is not carried past model validation.
D_soil_w    <- post_draws(gi_soil_w,   nd_soil)    # soil, relative scale
D_broth     <- post_draws(gi_broth,    nd_broth)   # broth, relative scale
D_broth_abs <- post_draws(m_broth_abs, nd_broth)   # broth, absolute (sensitivity)

dim(D_soil_w)
```

## Per-Draw Metrics

```{r}
# ---- Relative scale: the same three metrics for both media ----------------
rel_stats <- bind_rows(
  metric_set(D_soil_w, nd_soil,  "strain", THR_REL) %>% mutate(medium = "soil"),
  metric_set(D_broth,  nd_broth, "strain", THR_REL) %>% mutate(medium = "broth")
)

# ---- Absolute-scale broth, sensitivity only -------------------------------
abs_stats <- metric_set(D_broth_abs, nd_broth, "strain", THR_BROTH_ABS) %>%
  mutate(medium = "broth (absolute)")
```

## Metric Summaries

```{r}
## p_above is the posterior probability that the expected value is still above
## the threshold at day 100, i.e. that the population has NOT yet fallen by a
## full 4 logs (relative scale) or below 0 (absolute scale).
## Time-to-threshold is suppressed where fewer than P_CROSSED of draws cross.
metric_summary <- bind_rows(rel_stats, abs_stats) %>%
  group_by(medium, stat) %>%
  group_modify(~ summarise_stat(.x, "strain")) %>%
  ungroup() %>%
  mutate(reportable = !(grepl("^time to", stat) & p_crossed < P_CROSSED),
         across(c(median, q25, q75, lwr, upr),
                ~ ifelse(reportable, .x, NA_real_)))

print(metric_summary %>%
        select(medium, stat, strain, p_above, p_crossed, reportable,
               median, q25, q75, lwr, upr),
      n = Inf, width = Inf)

ggplot(filter(rel_stats, stat == "AUC"), aes(value)) +
  geom_histogram(bins = 60) +
  facet_grid(medium ~ strain, scales = "free",
             labeller = labeller(strain = strain_labs)) +
  labs(x = expression(AUC~above~-4~log*-fold~change~(log[10]%*%~days)), y = "Draws")
```

## Pairwise Contrasts

```{r}
rel_contrasts <- rel_stats %>%
  group_by(medium, stat) %>%
  group_modify(~ contrast_draws(.x, "strain", "x") %>% select(-stat)) %>%
  ungroup()

print(rel_contrasts %>% select(medium, stat, contrast, n_usable,
                               median, lwr, upr, p_gt0, excludes_0),
      n = Inf, width = Inf)
```

## Sensitivity: Broth on the Absolute Scale

```{r}
## Does modelling broth on the absolute scale change which strains differ?
## Contrasts are compared against the primary relative-scale broth model.
abs_contrasts <- abs_stats %>%
  group_by(medium, stat) %>%
  group_modify(~ contrast_draws(.x, "strain", "x") %>% select(-stat)) %>%
  ungroup()

print(abs_contrasts %>% select(stat, contrast, median, lwr, upr, excludes_0),
      n = Inf, width = Inf)

broth_sens <- bind_rows(lapply(unique(abs_contrasts$stat), function(st) {
  sens_agree(rel_contrasts %>% filter(medium == "broth", stat == st),
             abs_contrasts %>% filter(stat == st),
             st, c("relative", "absolute"))
}))

broth_sens %>% count(stat, agree)
broth_sens %>% filter(!agree) %>% print(n = Inf, width = Inf)
```

## Sensitivity: Common Floor Threshold (Soil)

```{r}
## Strains 1, 3 and 6 (and strain 2, by 0.03 log) have detection floors above
## -4 log, so part of the area, margin and crossing time measured against -4
## lies below anything the assay could detect for them. Here every strain is
## referenced to the highest soil detection floor (-3.59, strain 1): a level
## all six series could resolve directly, so the metrics stay comparable
## across strains. No refit; only the reference level changes.
soil_floors <- lod_rel %>%
  filter(medium == "soil", group == "exp") %>%
  arrange(strain) %>%
  transmute(strain, own_floor = y_lod)

THR_COMMON <- max(soil_floors$own_floor)

soil_floors %>% mutate(own_floor = round(own_floor, 3))
round(THR_COMMON, 3)

common_stats <- metric_set(D_soil_w, nd_soil, "strain", THR_COMMON,
                           ttt_label = "time to threshold")

common_summary <- common_stats %>%
  group_by(stat) %>%
  group_modify(~ summarise_stat(.x, "strain")) %>%
  ungroup() %>%
  mutate(reportable = !(grepl("^time to", stat) & p_crossed < P_CROSSED))

print(common_summary %>%
        select(stat, strain, p_above, p_crossed, reportable, median, lwr, upr),
      n = Inf, width = Inf)

common_contrasts <- common_stats %>%
  group_by(stat) %>%
  group_modify(~ contrast_draws(.x, "strain", "x") %>% select(-stat)) %>%
  ungroup()

## Primary (-4) soil contrasts, relabelled so the metric names line up
rel_soil <- rel_contrasts %>%
  filter(medium == "soil") %>%
  mutate(stat = sub("^time to.*", "time to threshold", stat))

common_sens <- sens_agree_all(rel_soil, common_contrasts, c("thr_minus4", "common"))

common_sens %>% count(stat, agree)

## Any contrast that changed, shown in full under both thresholds
flipped <- filter(common_sens, !agree) %>% select(stat, contrast)
bind_rows(
  semi_join(rel_soil,         flipped, by = c("stat", "contrast")) %>% mutate(threshold = "-4"),
  semi_join(common_contrasts, flipped, by = c("stat", "contrast")) %>% mutate(threshold = "common")
) %>%
  select(stat, contrast, threshold, n_usable, median, lwr, upr, p_gt0, excludes_0) %>%
  arrange(stat, contrast, threshold) %>%
  print(n = Inf, width = Inf)

## Supplementary-table strings for this sensitivity ("NA; x%" TTR format)
common_summary %>%
  mutate(cell = ifelse(grepl("^time to", stat),
                       fmt_ttr(median, lwr, upr, p_crossed, reportable),
                       fmt_est(median, lwr, upr))) %>%
  select(strain, stat, cell) %>%
  tidyr::pivot_wider(names_from = stat, values_from = cell)
```

# Constant-Temperature Controls

## Control Dataset

```{r}
set.seed(SEED)

ft_ctrl <- ft %>%
  filter(strain %in% c("4", "6")) %>%
  droplevels() %>%
  mutate(sg = interaction(strain, group, drop = TRUE, sep = "_"))

ctrl_soil  <- filter(ft_ctrl, medium == "soil")  %>% droplevels()
ctrl_broth <- filter(ft_ctrl, medium == "broth") %>% droplevels()

levels(ft_ctrl$sg)

ft_ctrl %>%
  group_by(medium, group) %>%
  summarise(n = n(), n_censored = sum(censored), .groups = "drop")

ctrl_broth %>%
  filter(time == 0) %>%
  group_by(sg) %>%
  summarise(baseline_log10 = round(mean(y_abs), 2), .groups = "drop")

## The regime comparisons made within each strain
ctrl_pairs <- list(
  c("4_exp",     "4_plus10C"),
  c("4_exp",     "4_minus10C"),
  c("4_plus10C", "4_minus10C"),
  c("6_exp",     "6_plus10C"),
  c("6_exp",     "6_minus10C"),
  c("6_plus10C", "6_minus10C")
)
```

## Models

```{r}
## Same GI structure as before, with the six-level strain x regime factor as
## the grouping variable.

## The soil fit below is the UNWEIGHTED model. It exists only to supply
## residuals to the scale model in the next chunk; m_ctrl_soil_w is what carries
## the control-arm inference, matching the treatment of gi_soil_w in the
## cycling arm (Variation 1).
m_ctrl_soil <- gam(
  cbind(y_obs, y_cens) ~ s(time, k = K_GLOBAL, m = 2, pc = 0) +
                         s(time, by = sg, k = K_GROUP, m = 1, pc = 0) - 1,
  data = ctrl_soil, family = cnorm(), method = "REML"
)

m_ctrl_broth <- gam(
  y_obs ~ s(time, k = K_GLOBAL, m = 2, pc = 0) +
          s(time, by = sg, k = K_GROUP, m = 1, pc = 0) - 1,
  data = ctrl_broth, family = gaussian(), method = "REML"
)

## Absolute scale
m_ctrl_broth_abs <- gam(
  y_abs ~ sg + s(time, k = K_GLOBAL, m = 2) + s(time, by = sg, k = K_GROUP, m = 1),
  data = ctrl_broth, family = gaussian(), method = "REML"
)

## Broth is left unweighted in both arms
lapply(list(broth_rel = m_ctrl_broth, broth_abs = m_ctrl_broth_abs), k.check)
```

## Variation: Control Soil w/ Time-Varying Residual Scale

```{r}
 # ---- Stage 1: regime-specific residual scale -------------------------------
## Pooling cycling and constant series flattens the variance ramp: the four
## constant-temperature series are tight throughout (0 of 156 non-detections)
## and dominate a single s(time). Fit the two regimes separately.
ctrl_soil_resid <- ctrl_soil %>%
  mutate(mu = as.vector(predict(m_ctrl_soil)), r = y_obs - mu) %>%
  filter(!censored) %>%
  mutate(regime = factor(ifelse(grepl("exp$", sg), "cycling", "constant"),
                         levels = c("cycling", "constant")))

table(ctrl_soil_resid$regime)

scale_mod_ctrl <- gam(I(r^2) ~ regime + s(time, k = 5, by = regime),
                      data = ctrl_soil_resid,
                      family = Gamma(link = "log"), method = "REML")
summary(scale_mod_ctrl)
k.check(scale_mod_ctrl)        # drop to k = 4 if the cycling smooth complains

# ---- Fitted scale, by regime ----------------------------------------------
## The grid needs a regime column now that the scale model has a by term.
scale_grid_ctrl <- expand.grid(
  time   = seq(0, TMAX, length.out = 200),
  regime = factor(c("cycling", "constant"), levels = c("cycling", "constant"))
)
scale_grid_ctrl$sd_hat <- sqrt(as.vector(
  predict(scale_mod_ctrl, newdata = scale_grid_ctrl, type = "response")))

ggplot(scale_grid_ctrl, aes(time, sd_hat, colour = regime)) +
  geom_point(data = ctrl_soil %>%
               mutate(regime = factor(ifelse(grepl("exp$", sg), "cycling", "constant"),
                                      levels = c("cycling", "constant"))) %>%
               group_by(sg, regime, time) %>%
               summarise(sd_y = sd(y_obs), .groups = "drop"),
             aes(time, sd_y, colour = regime), inherit.aes = FALSE,
             alpha = 0.55, size = 1.4) +
  geom_line(linewidth = 0.9) +
  geom_hline(yintercept = get_sigma(m_ctrl_soil), linetype = 2, colour = "grey40") +
  labs(x = "Day", y = "Residual SD (log10 units)", colour = NULL,
       subtitle = "Lines = fitted scale model; dashed = single sigma from the unweighted fit")

# ---- Stage 2: refit GI with inverse-variance prior weights -----------------
## regime must be present here too, for the predict() call below.
ctrl_soil_w <- ctrl_soil %>%
  mutate(regime = factor(ifelse(grepl("exp$", sg), "cycling", "constant"),
                         levels = c("cycling", "constant")))

ctrl_soil_w$var_hat <- as.vector(predict(scale_mod_ctrl, newdata = ctrl_soil_w,
                                         type = "response"))
ctrl_soil_w$w <- 1 / ctrl_soil_w$var_hat
ctrl_soil_w$w <- ctrl_soil_w$w / mean(ctrl_soil_w$w)

ctrl_soil_w %>%
  group_by(regime, time) %>%
  summarise(w = round(first(w), 3), sd_implied = round(sqrt(first(var_hat)), 3),
            .groups = "drop") %>%
  tidyr::pivot_wider(names_from = regime, values_from = c(w, sd_implied))

## THE MODEL ITSELF -- everything below depends on this line
m_ctrl_soil_w <- gam(
  cbind(y_obs, y_cens) ~ s(time, k = K_GLOBAL, m = 2, pc = 0) +
                         s(time, by = sg, k = K_GROUP, m = 1, pc = 0) - 1,
  data = ctrl_soil_w, family = cnorm(), weights = w, method = "REML"
)

tibble::tibble(
  fit   = c("control soil unweighted", "control soil weighted"),
  sigma = c(get_sigma(m_ctrl_soil), get_sigma(m_ctrl_soil_w)),
  edf   = c(sum(m_ctrl_soil$edf),   sum(m_ctrl_soil_w$edf))
)

k.check(m_ctrl_soil_w)

# ---- Did the weighting flatten the residual spread? ------------------------
ctrl_soil_w %>%
  mutate(mu = as.vector(predict(m_ctrl_soil_w)), r = y_obs - mu,
         rstd = r * sqrt(w)) %>%
  filter(!censored) %>%
  ggplot(aes(time, rstd)) +
  geom_hline(yintercept = 0, linetype = 2, colour = "grey60") +
  geom_point(aes(colour = sg), alpha = 0.65, size = 1.3) +
  geom_smooth(aes(y = abs(rstd)), method = "loess", se = FALSE,
              colour = "firebrick", linewidth = 0.7) +
  labs(x = "Day", y = "Weighted residual", colour = NULL,
       subtitle = "Red = loess of |residual|; should now be roughly flat")

# ---- Detection scores, weighted vs unweighted ------------------------------
## Prior weights scale the per-observation SD as sigma / sqrt(w).
detect_check_ctrl <- function(m, label, wts = NULL) {
  sd_i <- if (is.null(wts)) get_sigma(m) else get_sigma(m) / sqrt(wts)
  ctrl_soil %>%
    mutate(model   = label,
           mu      = as.vector(predict(m, newdata = ctrl_soil)),
           y_lod   = log10(lod) - log10_baseline,
           p_below = pnorm(y_lod, mean = mu, sd = sd_i))
}

dc_ctrl <- bind_rows(
  detect_check_ctrl(m_ctrl_soil,   "unweighted"),
  detect_check_ctrl(m_ctrl_soil_w, "weighted", wts = ctrl_soil_w$w)
)

score_detect(dc_ctrl, "control soil")

dc_ctrl %>%
  group_by(sg, model) %>%
  summarise(observed = sum(censored),
            expected = round(sum(p_below), 1), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = model, values_from = expected,
                     names_prefix = "expected_", id_cols = c(sg, observed))
```


## Posterior Draws and Metrics

```{r}
nd_ctrl_soil  <- expand.grid(time = t_grid,
                             sg = factor(levels(ctrl_soil_w$sg),
                                         levels = levels(ctrl_soil_w$sg)))
nd_ctrl_broth <- expand.grid(time = t_grid,
                             sg = factor(levels(ctrl_broth$sg),
                                         levels = levels(ctrl_broth$sg)))

Dc_soil      <- post_draws(m_ctrl_soil_w,   nd_ctrl_soil)
Dc_broth     <- post_draws(m_ctrl_broth,    nd_ctrl_broth)
Dc_broth_abs <- post_draws(m_ctrl_broth_abs, nd_ctrl_broth)

## Same three metrics as the cycling arm, same thresholds.
ctrl_stats <- bind_rows(
  metric_set(Dc_soil,  nd_ctrl_soil,  "sg", THR_REL) %>% mutate(medium = "soil"),
  metric_set(Dc_broth, nd_ctrl_broth, "sg", THR_REL) %>% mutate(medium = "broth")
)

## Absolute-scale broth, sensitivity only
ctrl_abs_stats <- metric_set(Dc_broth_abs, nd_ctrl_broth, "sg", THR_BROTH_ABS) %>%
  mutate(medium = "broth (absolute)")

ctrl_summary <- bind_rows(ctrl_stats, ctrl_abs_stats) %>%
  group_by(medium, stat) %>%
  group_modify(~ summarise_stat(.x, "sg")) %>%
  ungroup() %>%
  mutate(reportable = !(grepl("^time to", stat) & p_crossed < P_CROSSED),
         across(c(median, q25, q75, lwr, upr),
                ~ ifelse(reportable, .x, NA_real_)))

print(ctrl_summary %>% select(medium, stat, sg, p_above, p_crossed, reportable,
                              median, q25, q75, lwr, upr),
      n = Inf, width = Inf)
```

## Within-Strain Regime Contrasts

```{r}
ctrl_contrasts <- ctrl_stats %>%
  group_by(medium, stat) %>%
  group_modify(~ contrast_draws(.x, "sg", "x", ctrl_pairs) %>% select(-stat)) %>%
  ungroup()

print(ctrl_contrasts %>% select(medium, stat, contrast, n_usable,
                                median, lwr, upr, p_gt0, excludes_0),
      n = Inf, width = Inf)

## Absolute-scale broth sensitivity: same regime comparisons, compared against
## the primary relative-scale broth model.
ctrl_abs_contrasts <- ctrl_abs_stats %>%
  group_by(medium, stat) %>%
  group_modify(~ contrast_draws(.x, "sg", "x", ctrl_pairs) %>% select(-stat)) %>%
  ungroup()

print(ctrl_abs_contrasts %>% select(stat, contrast, median, lwr, upr, excludes_0),
      n = Inf, width = Inf)

bind_rows(lapply(unique(ctrl_abs_contrasts$stat), function(st) {
  sens_agree(ctrl_contrasts     %>% filter(medium == "broth", stat == st),
             ctrl_abs_contrasts %>% filter(stat == st),
             st, c("relative", "absolute"))
})) %>% count(stat, agree)
```

## Control Diagnostic Plots

```{r}
bind_rows(
  pred_grid(m_ctrl_soil_w, nd_ctrl_soil,  "soil (relative)",  tag = "panel"),
  pred_grid(m_ctrl_broth,  nd_ctrl_broth, "broth (relative)", tag = "panel")
) %>%
  tidyr::separate(sg, into = c("strain", "group"), sep = "_", remove = FALSE) %>%
  mutate(group  = factor(group, levels = c("exp", "plus10C", "minus10C")),
         strain = factor(strain, levels = levels(ft_ctrl$strain))) %>%
  ggplot(aes(time, fit, colour = group, fill = group)) +
  geom_hline(yintercept = 0, linetype = 2, colour = "grey60") +
  geom_ribbon(aes(ymin = fit - 2 * se, ymax = fit + 2 * se),
              alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.8) +
  geom_point(data = ft_ctrl, aes(time, y_obs, colour = group),
             inherit.aes = FALSE, alpha = 0.3, size = 1) +
  facet_grid(panel ~ strain, scales = "free_y",
             labeller = labeller(strain = strain_labs)) +
  labs(x = "Day", y = expression(log[10]~fold~change~from~baseline),
       colour = NULL, fill = NULL) +
  theme(legend.position = "bottom")

pred_grid(m_ctrl_broth_abs, nd_ctrl_broth, "broth (absolute)", tag = "panel") %>%
  tidyr::separate(sg, into = c("strain", "group"), sep = "_", remove = FALSE) %>%
  mutate(group  = factor(group, levels = c("exp", "plus10C", "minus10C")),
         strain = factor(strain, levels = levels(ft_ctrl$strain))) %>%
  ggplot(aes(time, fit, colour = group, fill = group)) +
  geom_ribbon(aes(ymin = fit - 2 * se, ymax = fit + 2 * se),
              alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.8) +
  geom_point(data = ctrl_broth, aes(time, y_abs, colour = group),
             inherit.aes = FALSE, alpha = 0.3, size = 1) +
  facet_wrap(~ strain, labeller = labeller(strain = strain_labs)) +
  labs(x = "Day", y = expression(log[10]~CFU~mL^-1),
       colour = NULL, fill = NULL, title = "Broth on the absolute scale") +
  theme(legend.position = "bottom")
```

# Figures

## Style

```{r}
lw <- function(pt) pt / .pt

STRAIN_COL <- c(`1` = "#E69F00", `2` = "#56B4E9", `3` = "#009E73",
                `4` = "#0072B2", `5` = "#D55E00", `6` = "#CC79A7")

STRAIN_LTY <- c(`1` = "42",   `2` = "22",   `3` = "1343",
                `4` = "44",   `5` = "solid", `6` = "73")

STRAIN_LWD <- c(`1` = lw(1), `2` = lw(1), `3` = lw(1),
                `4` = lw(1), `5` = lw(1), `6` = lw(1))

FIG_LABS <- expression(
  "1  Poultry, Belgium (clade 2)",
  "2  Arctic fox, Victoria Is. (clade 2)",
  "3  Swine, Belgium (clade 3)",
  "4  Muskox, Ellesmere Is. (clade 3)",
  "5  Muskox, Ellesmere Is. (clade 3, Arctic clone)",
  "6  " * italic("E. qupirruqcola") * ", Ellesmere Is."
)

PANEL_LABS <- c(broth = "(a)  Broth", soil = "(b)  Soil")

## Legend fills down the first column (1,2,3) then the second (4,5,6)
strain_guide <- guide_legend(ncol = 2, byrow = FALSE)

## 4 log10 reduction reference line
THR_COL <- "grey10"
THR_LWD <- lw(2.0)

Y_FLOOR_SOIL <- -5      # soil ribbons squished to this floor, not dropped

theme_fig <- theme_bw(base_size = 8, base_family = "") +
  theme(
    panel.grid.minor  = element_blank(),
    panel.grid.major  = element_line(linewidth = lw(0.25), colour = "grey88"),
    panel.border      = element_rect(linewidth = lw(0.5), colour = "grey20"),
    axis.text         = element_text(size = 7, colour = "black"),
    axis.title        = element_text(size = 8),
    axis.ticks        = element_line(linewidth = lw(0.5), colour = "grey20"),
    strip.background  = element_blank(),
    strip.text        = element_text(size = 8, hjust = 0, face = "bold",
                                     margin = margin(b = 3)),
    legend.text       = element_text(size = 7),
    legend.title      = element_blank(),
    legend.key.width  = unit(11, "mm"),
    legend.key.height = unit(3.5, "mm"),
    legend.margin     = margin(t = 0),
    legend.position   = "bottom",
    plot.margin       = margin(2, 4, 2, 2)
  )
```

## Fig. 1

```{r}
## Median and 95% interval of the posterior at every grid point.
ribbon_from_draws <- function(D, nd, medium_label) {
  q <- t(apply(D, 1, quantile, probs = c(0.025, 0.5, 0.975)))
  tibble::tibble(
    medium = factor(medium_label, levels = levels(ft$medium)),
    time   = nd$time,
    strain = nd$strain,
    lwr    = q[, 1],
    med    = q[, 2],
    upr    = q[, 3]
  )
}

## Cut a series where its posterior median falls below `thr` for the rest of
## the trial. Series whose median is at or above
## thr on the last grid day are returned whole, including any earlier dips
## below thr. The cut point is interpolated to sit exactly on thr.
## cut_day records where each series was cut (NA = shown in full).
cut_at_final_crossing <- function(d, gvar, thr = THR_REL) {
  d %>%
    group_by(across(all_of(gvar))) %>%
    group_modify(function(g, key) {
      g <- arrange(g, time)
      n <- nrow(g)
      if (g$med[n] >= thr) return(mutate(g, cut_day = NA_real_))
      k <- max(which(g$med >= thr))               # last grid point at/above thr
      f <- (thr - g$med[k]) / (g$med[k + 1] - g$med[k])
      cross <- g[k, ]
      cross$time <- g$time[k] + f * (g$time[k + 1] - g$time[k])
      cross$lwr  <- g$lwr[k]  + f * (g$lwr[k + 1]  - g$lwr[k])
      cross$upr  <- g$upr[k]  + f * (g$upr[k + 1]  - g$upr[k])
      cross$med  <- thr
      bind_rows(g[seq_len(k), ], cross) %>% mutate(cut_day = cross$time)
    }) %>%
    ungroup()
}

## Display-only: hide any stretch where the median runs below the plotting
## floor (so lines end at the floor instead of being flattened along it).
## Each visible stretch gets its own seg_id for the `group` aesthetic.
mask_below_threshold <- function(d, gvar, thr = THR_REL) {
  d %>%
    group_by(across(all_of(gvar))) %>%
    group_modify(function(g, key) {
      g <- arrange(g, time)
      n <- nrow(g)
      above <- g$med >= thr
      i <- which(above[-n] != above[-1])          # crossings between grid points
      if (length(i) > 0) {
        f <- (thr - g$med[i]) / (g$med[i + 1] - g$med[i])
        cross <- g[i, ]
        cross$time <- g$time[i] + f * (g$time[i + 1] - g$time[i])
        cross$lwr  <- g$lwr[i]  + f * (g$lwr[i + 1]  - g$lwr[i])
        cross$upr  <- g$upr[i]  + f * (g$upr[i + 1]  - g$upr[i])
        cross$med  <- thr
        g <- bind_rows(g, cross) %>% arrange(time)
      }
      keep  <- g$med >= thr - 1e-9
      g$seg <- cumsum(keep & !dplyr::lag(keep, default = FALSE))
      g[keep, ]
    }) %>%
    ungroup() %>%
    mutate(seg_id = paste(.data[[gvar]], seg, sep = "_"))
}

fig1_broth <- ribbon_from_draws(D_broth, nd_broth, "broth") %>%
  mutate(seg = 1L, seg_id = paste(strain, seg, sep = "_"))

fig1_soil <- ribbon_from_draws(D_soil_w, nd_soil, "soil") %>%
  cut_at_final_crossing("strain", thr = THR_REL) %>%
  mask_below_threshold("strain", thr = Y_FLOOR_SOIL) %>%
  ## Lower ribbon bounds can still hang below the floor; squish those only
  mutate(lwr = pmax(lwr, Y_FLOOR_SOIL))

fig1_dat <- bind_rows(fig1_broth, fig1_soil)

## Which soil curves were cut, and where. cut_day = NA means shown in full.
## A to_day short of 100 for an uncut strain means its median passes
## Y_FLOOR_SOIL somewhere
fig1_soil %>%
  group_by(strain, seg) %>%
  summarise(from_day = round(min(time), 1), to_day = round(max(time), 1),
            min_med  = round(min(med), 2),
            cut_day  = round(first(cut_day), 1), .groups = "drop")

## Threshold line belongs only in the soil panel
thr_line <- tibble::tibble(
  medium = factor("soil", levels = levels(ft$medium)),
  y      = THR_REL
)

fig1 <- ggplot(fig1_dat, aes(time, med, colour = strain, fill = strain)) +
  geom_hline(yintercept = 0, linetype = "dotted",
             linewidth = lw(0.4), colour = "grey45") +
  geom_ribbon(aes(ymin = lwr, ymax = upr, group = seg_id),
              alpha = 0.13, colour = NA) +
  geom_hline(data = thr_line, aes(yintercept = y), inherit.aes = FALSE,
             linewidth = THR_LWD, colour = THR_COL) +
  geom_line(aes(linetype = strain, linewidth = strain, group = seg_id)) +
  scale_colour_manual(values = STRAIN_COL, labels = FIG_LABS, name = NULL) +
  scale_fill_manual(values = STRAIN_COL,   labels = FIG_LABS, name = NULL) +
  scale_linetype_manual(values = STRAIN_LTY, labels = FIG_LABS, name = NULL) +
  scale_linewidth_manual(values = STRAIN_LWD, labels = FIG_LABS, name = NULL) +
  scale_x_continuous(breaks = seq(0, TMAX, 20), expand = expansion(mult = 0.01)) +
  facet_wrap(~ medium, ncol = 1, scales = "free_y",
             labeller = labeller(medium = PANEL_LABS)) +
  labs(x = "Day (one freeze/thaw cycle per day)",
       y = expression(log[10]~fold~change~from~baseline)) +
  guides(colour = strain_guide, linetype = strain_guide,
         linewidth = strain_guide, fill = "none") +
  theme_fig

fig1

FIG_W <- 170   # mm, full two-column width
FIG_H <- 145   # mm

ggsave("Figure_1.pdf", fig1, width = FIG_W, height = FIG_H, units = "mm",
       device = cairo_pdf)

ggsave("Figure_1.png", fig1, width = FIG_W, height = FIG_H, units = "mm",
       dpi = 600, device = ragg::agg_png)
```
## Fig. 2

```{r}
## Modeled trajectories for strains 4 and 6 under freeze/thaw cycling,
## constant +10 C and constant -10 C, in (a) broth and (b) soil.
## Draws come from the control models (m_ctrl_broth, m_ctrl_soil_w).
## Color = strain; line type, opacity and width = temperature regime.
## Soil series use the same display rule as Fig. 1.

REGIME_LEVS  <- c("exp", "plus10C", "minus10C")
REGIME_LABS  <- c(exp      = "Freeze/Thaw Cycling",
                  plus10C  = "Constant +10\u00B0C",
                  minus10C = "Constant \u221210\u00B0C")
REGIME_LTY   <- c(exp = "solid", plus10C = "44",   minus10C = "13")
REGIME_ALPHA <- c(exp = 1,       plus10C = 0.65,   minus10C = 0.65)
REGIME_LWD   <- c(exp = lw(1.5), plus10C = lw(1.3), minus10C = lw(1.1))

RIBBON_ALPHA   <- 0.10
RIBBON_REGIMES <- REGIME_LEVS   # e.g. c("exp") to ribbon cycling only

## Median and 95% interval at every grid point, split into strain and regime.
ribbon_from_draws_sg <- function(D, nd, medium_label) {
  q <- t(apply(D, 1, quantile, probs = c(0.025, 0.5, 0.975)))
  tibble::tibble(
    medium = factor(medium_label, levels = levels(ft$medium)),
    time   = nd$time,
    sg     = nd$sg,
    lwr    = q[, 1],
    med    = q[, 2],
    upr    = q[, 3]
  ) %>%
    tidyr::separate(sg, into = c("strain", "group"), sep = "_", remove = FALSE) %>%
    mutate(strain = factor(strain, levels = levels(ft$strain)),
           group  = factor(group,  levels = REGIME_LEVS))
}

fig2_broth <- ribbon_from_draws_sg(Dc_broth, nd_ctrl_broth, "broth") %>%
  mutate(seg = 1L, seg_id = paste(sg, seg, sep = "_"))

fig2_soil <- ribbon_from_draws_sg(Dc_soil, nd_ctrl_soil, "soil") %>%
  cut_at_final_crossing("sg", thr = THR_REL) %>%
  mask_below_threshold("sg", thr = Y_FLOOR_SOIL) %>%
  mutate(lwr = pmax(lwr, Y_FLOOR_SOIL))

fig2_dat <- bind_rows(fig2_broth, fig2_soil)

## Which soil series were cut, and where (cut_day = NA means shown in full)
fig2_soil %>%
  group_by(sg, seg) %>%
  summarise(from_day = round(min(time), 1), to_day = round(max(time), 1),
            min_med  = round(min(med), 2),
            cut_day  = round(first(cut_day), 1), .groups = "drop")

fig2 <- ggplot(fig2_dat, aes(time, med)) +
  geom_hline(yintercept = 0, linetype = "dotted",
             linewidth = lw(0.4), colour = "grey45") +
  geom_ribbon(data = filter(fig2_dat, group %in% RIBBON_REGIMES),
              aes(ymin = lwr, ymax = upr, fill = strain, group = seg_id),
              alpha = RIBBON_ALPHA, colour = NA) +
  geom_hline(data = thr_line, aes(yintercept = y), inherit.aes = FALSE,
             linewidth = THR_LWD, colour = THR_COL) +
  geom_line(aes(colour = strain, linetype = group, alpha = group,
                linewidth = group, group = seg_id)) +
  scale_colour_manual(values = STRAIN_COL, breaks = c("4", "6"),
                      labels = FIG_LABS[c(4, 6)], name = NULL) +
  scale_fill_manual(values = STRAIN_COL, guide = "none") +
  scale_linetype_manual(values = REGIME_LTY,   labels = REGIME_LABS,
                        breaks = REGIME_LEVS, name = NULL) +
  scale_alpha_manual(values = REGIME_ALPHA,    labels = REGIME_LABS,
                     breaks = REGIME_LEVS, name = NULL) +
  scale_linewidth_manual(values = REGIME_LWD,  labels = REGIME_LABS,
                         breaks = REGIME_LEVS, name = NULL) +
  scale_x_continuous(breaks = seq(0, TMAX, 20), expand = expansion(mult = 0.01)) +
  facet_wrap(~ medium, ncol = 1, scales = "free_y",
             labeller = labeller(medium = PANEL_LABS)) +
  labs(x = "Day",
       y = expression(log[10]~fold~change~from~baseline)) +
  guides(
    colour    = guide_legend(order = 1, ncol = 1,
                             override.aes = list(linetype = "solid", alpha = 1,
                                                 linewidth = lw(1.1))),
    linetype  = guide_legend(order = 2, ncol = 1,
                             override.aes = list(colour = "grey20")),
    alpha     = guide_legend(order = 2, ncol = 1),
    linewidth = guide_legend(order = 2, ncol = 1)
  ) +
  theme_fig

fig2

FIG2_W <- 170   # mm, full two-column width
FIG2_H <- 145   # mm

ggsave("Figure_2.pdf", fig2, width = FIG2_W, height = FIG2_H, units = "mm",
       device = cairo_pdf)

ggsave("Figure_2.png", fig2, width = FIG2_W, height = FIG2_H, units = "mm",
       dpi = 600, device = ragg::agg_png)
```

# Session Information

```{r}
sessionInfo()
beep(sound = "fanfare")
```
